---
title: Documentation `su2` project
author: Simone Romiti
output: 
  html_document:
    toc: true
    theme: united
bibliography: biblio.bib  
csl: 2d-materials.csl
link-citations: true
---

The present repository contains routines for numerical simulations of $SU(2)$ and $U(1)$ lattice gauge theories in $d=2,3,4$ dimensions.

# Library routines

The gauge configurations can be generated with the following Markov Chain Monte Carlo algorithms (see @gattringer2009quantum for the theoretical background):

- Metropolis
- Hybrid Monte Carlo

The measurements can be generated either online (during the MCMC) or offline (on previously generated configurations). At present the supported observables are:

- Plaquette standard and clover-improved energy
- Glueball correlators
- "pion" correlator
- Wilson loops and static potential among $\bar{q} q$ pairs
- Wilson gradient flow

### Programs info

- Programs: `main-u1.cpp` and `main-su2.cpp`
- Parameters parsing: `yaml` input file
- Input file example: see [./main.input](main.input)
  Notes:
  - Each run produces/updates a file named `nconf_counter.txt` in the directory storing the gauge configurations. This file contains a header line and the values below. An example:
  ```
  heat i path_conf
  1 1140 ./confs/config_u1.4.4.1.16.b1.900000.1140
  ```
  - `restart` and `heat` cannot be passed simultaneously
  - If `restart: true` is passed, the last configuration id and path are read from the `nconf_counter.txt` file.
  - The program can do online and offline measurements

##### `hadron` folder 

In the `hadron/` folder are found scripts for data manipulation such that online measurements are more easily readable with the `R` library `hadron`: https://github.com/HISKP-LQCD/hadron. See the `README.md` in the `hadron/` folder for more information.


### tests

In the `test/` folder the user can find various simple programs implementing the routines of this repository. For further details please look at the `test/README.md` file.

# How to use the library

```{r child = './howto.Rmd'}
```

# Theoretical background

Fermions are included as a degenerate doublet using the staggered discretization.

```{r child = './staggered_fermions.Rmd'}
```
